Convergence rate of MAP estimates \\ for the exponential family

### 2021-03-29 main.tex
Bias Variance discussion

\section{Asymptotic Behaviour}
In the limit, the MAP estimates reach a rate $O(1/2n)$. This is shown by approximating the Bregman divergence using a second order Taylor expansion
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    &= \half (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
    + O(\norm{\mu - \mu^*}^3)\\
    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
    + O(\norm{\mu - \mu^*}^3)
\end{align}
where we introduced the Mahalanobis distance induced by the matrix 
\begin{align}
    \nabla^2\conj(\mu^*) 
    = \nabla^2\logpart(\natp^*)^{-1} 
    = \Cov_{\natp^*}[T(X)]^{-1} 
    =: \mSigma^{-1}  \; .
\end{align}
To exploit this approximation, let us extend the $\ell^2$ norm results to a Mahalanobis distance induced by matrix $\mM$
\begin{align}
	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
\end{align}
where $\mSigma$ is the covariance of $T(X)$. We retrieve $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$.  For the MLE we get
\begin{align}
	\E \bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) = \frac{d}{2n} + O(n^{- \frac{3}{2}})
\end{align}
and for the MAP we  get 
\begin{align}
	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
	&= \frac{n d}{(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
	&= O\left(\frac{d}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
\end{align}
Remark how the variance does not even appear, only the dimension divided by $n$ really matters. I find this result quite spectacular : the convergence speed of MLE is not affected by the covariance of sufficient statistics, and for MAP it matters only in the $O(n^{-2})$ term.
Indeed this is because it is hidden within the Bregman divergence loss. 
Anyway that's all good for asymptotic results, but we are interested in a finite sample analysis. How do we get this ?

\section{Bias-Variance Decomposition}
Using Jensen inequality, we have the following analog of the bias-variance decomposition:

@@

\begin{align}
	& \expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
	= \conj(\mu^* ) 
	- \underbrace{\expect{\conj(\hat \mu_n)}}_{\leq \conj(\mu_n) \text{ (Jensen)} } 
	\pm \langle \nabla \conj( \mu_n) ; \mu^* - \mu_n \rangle
	- \E[\langle \hat \natp_n ; \mu^* - \hat \mu_n  \rangle ] \\
	& \leq \conj(\mu^* ) 
	- \conj(\mu_n) 
	- \langle \nabla \conj( \mu_n) ; \mu^* - \mu_n \rangle
	+ \frac{n_0}{n+n_0} \langle \natp_n ; \mu^* - \mu_0 \rangle \nonumber\\
	& \quad \quad \quad 
	- \frac{n_0}{n+n_0} \langle \E[\hat \natp_n] ; \mu^*- \mu_0  \rangle
	- \inv{n+n_0} \E[\langle \hat \natp_n ; \mu^* - T(X_i)  \rangle]\\
	& \leq \bias{\bregmanconj(\mu^*; \mu_n)}
	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_* - \mu_0} ; \variance{\natp_n - \E[\hat \natp_n]} \rangle 
	+ \frac{1}{n+n_0} \smallsum_i \variance{\expect{\langle \hat \natp_n ; T(X_i) - \mu^* \rangle } }
\end{align}
where $\hat \mu_n$ is the mean MAP estimate, $\hat \natp_n = \nabla \conj(\hat \mu_n)$ is the natural MAP estimate, 
$\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$ is the expected MAP estimate,
and $\theta_n = \nabla \conj(\mu_n)$ is the natural parameter of this expectation.

Note that unless $\conj$ is $\ell^2$, or degenerate 0 variance cases, $\theta_n \neq \expect{\hat \theta_n}$.
In fact, $\theta_n - \expect{\hat \theta_n} = \nabla \conj(\E[\hat \mu_n]) - \E[\nabla \conj(\hat \mu_n)]$

>>>>

\section{Asymptotic Behaviour}
In the limit, the MAP estimates reach a rate $O(1/2n)$. This is shown by approximating the Bregman divergence using a second order Taylor expansion
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    &= \half (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
    + O(\norm{\mu - \mu^*}^3)\\
    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
    + O(\norm{\mu - \mu^*}^3)
\end{align}
where we introduced the Mahalanobis distance induced by the matrix 
\begin{align}
    \nabla^2\conj(\mu^*) 
    = \nabla^2\logpart(\natp^*)^{-1} 
    = \Cov_{\natp^*}[T(X)]^{-1} 
    =: \mSigma^{-1}  \; .
\end{align}
To exploit this approximation, let us extend the $\ell^2$ norm results to a Mahalanobis distance induced by matrix $\mM$
\begin{align}
	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
\end{align}
where $\mSigma$ is the covariance of $T(X)$. We retrieve $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$.  For the MLE we get
\begin{align}
	\E \bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) = \frac{d}{2n} + O(n^{- \frac{3}{2}})
\end{align}
and for the MAP we  get 
\begin{align}
	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
	&= \frac{n d}{(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
	&= \frac{d}{n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
\end{align}
or 
\begin{align}
	\E \bregmanconj \left (\E [T(X)] ;  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) = \frac{d}{2n} + O(n^{- \frac{3}{2}})
\end{align}
Remark how the variance does not even appear, only the dimension divided by $n$ really matters. I find this result quite spectacular : the convergence speed of MLE is not affected by the covariance of sufficient statistics, and for MAP it matters only in the $O(n^{-2})$ term.
Indeed this is because it is hidden within the Bregman divergence loss. 
Anyway that's all good for asymptotic results, but we are interested in a finite sample analysis. How do we get this ?

\section{Bias-Variance Decomposition}
Using Jensen inequality and inserting the right quantities, we obtain the following analog of the bias-variance decomposition:

@@

\begin{align}
	& \expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
	= \conj(\mu^* ) 
	- \underbrace{\expect{\conj(\hat \mu_n)}}_{\leq \conj(\mu_n) \text{ (Jensen)} } 
	\pm \langle \nabla \conj( \mu_n) ; \mu^* - \mu_n \rangle
	- \E[\langle \hat \natp_n ; \mu^* - \hat \mu_n  \rangle ] \\
	& \leq \conj(\mu^* ) 
	- \conj(\mu_n) 
	- \langle \nabla \conj( \mu_n) ; \mu^* - \mu_n \rangle
	+ \frac{n_0}{n+n_0} \langle \natp_n ; \mu^* - \mu_0 \rangle \nonumber\\
	& \quad \quad \quad 
	- \frac{n_0}{n+n_0} \langle \E[\hat \natp_n] ; \mu^*- \mu_0  \rangle
	- \inv{n+n_0} \smallsum_i \E[\langle \hat \natp_n ; \mu^* - T(X_i)  \rangle]\\
	& \leq \bias{\bregmanconj(\mu^*; \mu_n)}
	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_* - \mu_0} ; \variance{\natp_n - \E[\hat \natp_n]} \rangle 
	+ \frac{n}{n+n_0} \variance{\expect{\langle \hat \natp_n  ; T(X_0) - \mu^* \rangle } }
\end{align}
where $\hat \mu_n$ is the mean MAP estimate, $\hat \natp_n = \nabla \conj(\hat \mu_n)$ is the natural MAP estimate, 
$\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$ is the expected MAP estimate,
and $\theta_n = \nabla \conj(\mu_n)$ is the natural parameter of this expectation.

The interpretation of the bias term should be clear.
The variance term is called as such because it is the covariance between $\hat \natp_n$ and $T(X)$
\begin{align}
	\Cov \bigg(T(X_0), \nabla \conj \big(\frac{\smallsum_i T(X_i) +n_0 \mu_0}{n+n_0} \big)\bigg	) \; .
\end{align}
The mixed bias variance term is might be negative. 
\TODO{add figure.}
Note that unless $\conj$ is $\ell^2$, or degenerate 0 variance cases, $\theta_n \neq \expect{\hat \theta_n}$.
In fact, $\theta_n - \expect{\hat \theta_n} = \nabla \conj(\E[\hat \mu_n]) - \E[\nabla \conj(\hat \mu_n)]$ is sort of a commuting bracket.

\subsection{Asymptotes}
Using the quadratic approximation of the Bregman divergence near $\mu^*$, we get a $O(n^{-2})$ rate for the pure bias term
\begin{align}
	\bregmanconj(\mu^*; \mu_n) 
	= \frac{\norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}}}{2(1 + \frac{n}{n_0})^{-2} } + O(n^{-3}) \; .
\end{align}
Using a linear approximation of the mirror map, we also get a $O(n^{-1})$ rate for the variance term
\begin{align}
	&\nabla \conj(\mu)  
	= \nabla \conj(\mu^*) + \nabla^2\conj(\mu^*)(\mu - \mu^*) + O(\norm{\mu - \mu^*}^2) \\
	\implies &\hat \natp_n 
	= \natp^* + \Sigma^{-1} (\hat \mu_n - \mu^*)  O(\norm{\hat \mu_n - \mu^*}^2) \\
	\implies &\Cov(\hat \natp_n ; T(X) ) 
	= \inv{n+n_0} \E[ \norm{T(X) - \mu^*}^2_{\Sigma^{-1}}]   + O(n^{-2}) \\
	\implies &\Cov(\hat \natp_n ; T(X) ) 
	 = \frac{d}{n+n_0} + O(n^{-2}) \; .
\end{align}
where we lost the factor $2$ compared to the straightforward asymptotic analysis, probably because of the Jensen upper inequality.

Finally, we can use this same linear approximation to bound the mixed bias-variance term
